Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/abstract.tex

Staged program languages assign a stage to each expression and
evaluate each expression in its assigned stage.  For example, in a
two-staged program, some expressions are evaluated in the first stage
and others in the second stage.  Programming languages, evaluation
strategies, and type systems for staged programs where stages roughly
correspond to the arrival order of inputs have been studied
extensively in the areas of partial evaluation and meta programming.

In this paper, we consider staged programs written in a typed lambda
calculus with a (``circle'') modality denoting computation in a later
stages and present techniques for splitting them into multi-pass
programs. Such multi-pass programs evaluate in a sequence of
evaluations of un-staged, conventional programs each of which
corresponds to a pass.  Conceived in the 80`s but J{\o}rring and
Scherlis, this {\em pass-separation} technique has been studied but
existing techniques are either manual or apply only to simple programs
that lack important features such as recursion and first-class
functions.

>>>>

Staged programming languages assign a stage to each expression and
evaluate each expression in its assigned stage.  For example, in a
two-staged program, some expressions are evaluated in the first stage
and others in the second stage.  Programming languages, evaluation
strategies, and type systems for staged programs where stages roughly
correspond to the arrival order of inputs have been studied
extensively in the areas of partial evaluation and meta programming.

In this paper, we consider staged programs written in a typed lambda
calculus with a (``circle'') modality denoting computation in a later
stages and present techniques for splitting them into multi-pass
programs. Such multi-pass programs evaluate in a sequence of
evaluations of un-staged, conventional programs each of which
corresponds to a pass.  Conceived in the 80's but J{\o}rring and
Scherlis, this {\em pass-separation} technique has been studied but
existing techniques are either manual or apply only to simple programs
that lack important features such as recursion and first-class
functions.

### 2015-02-27 paper/intro.tex

\section{Introduction}

@@

A closely related technique, {\em metaprogramming}, allows writing programs that
generate specialized code. Metaprogramming enables fine-grained control over
specialization by requiring staging annotations that make explicit the stage of
each computation. (In partial evaluation, we typically assume that a
\emph{binding-time analysis} has automatically determined which parts of $f$ can
be specialized.)

>>>>

\section{Introduction}
\label{sec:intro}

@@

A closely related technique, {\em metaprogramming}, allows writing
programs that generate specialized code
(e.g.,~\cite{davies96,Taha97,DP01-modal,NP05-nn}. Metaprogramming
enables fine-grained control over specialization by requiring staging
annotations that make explicit the stage of each computation. (In
partial evaluation, we typically assume that a \emph{binding-time
  analysis} has automatically determined which parts of $f$ can be
specialized.)
